Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-28 overleaf/paper_SIMD.tex
Add translation to MOTION to paper

\subsection{From (Optimized) MPC Source to MOTION}
\label{sec:MOTION}
\ana{Ben, add the MOTION section here.}

>>>>

\subsection{From (Optimized) MPC Source to MOTION}
\label{sec:MOTION}

MOTION supports FOR loops and SIMD operations, so translation from MPC source to MOTION C++ code is relatively straightforward.
% Variable naming
Our generated C++ uses the following variable-naming scheme: shared variables are named the same as in the MPC source with the \texttt{!} replaced with an underscore (e.g. \texttt{sum!2} would be translated to \texttt{sum\_2}).  Plaintext variables follow the same naming convention as shared variables but are prefixed with \texttt{\_MPC\_PLAINTEXT\_}.  The shared representation of constants are named \texttt{\_MPC\_CONSTANT\_} followed by the literal constant (e.g. the shared constant \texttt{0} would be named \texttt{\_MPC\_CONSTANT\_0}).

% Function preamble
The generated MOTION code begins with the declaration of all variables used in the function, including loop counters.  If a variable is a vectorized array, it is initialized to a correctly-sized array of empty MOTION shares.  Additionally, each plaintext variable and parameter has a shared counterpart declared.  Next, all constant values which are used as part of shared expressions are initialized as a shared input from party 0.  Finally, all plaintext parameters are converted used as shared inputs from party 0 to initialize their shared counterparts.

% Non-vectorized assignments and returns
Once the function preamble is complete, the MPC source is translated into C++ one statement at a time.  If there is no vectorization present in a statement, translation to C++ is straightforward: outside of Mux statements and array updates, non-vectorized assignments, expressions, and returns directly translated into their C++ equivalents.  Unvectorized Mux statements are converted to MOTION's Mux member function on the condition variable.  Array updates are translated into two C++ assignments: one to update the value in the original array and one to assign the new array.

\begin{table*}
\begin{tabular}{ccc}
\begin{minipage}{0.33\textwidth}
{\small
\begin{pythonn}
A[i] = val
\end{pythonn}
}
\end{minipage}

\begin{minipage}{0.33\textwidth}
{\small
\begin{pythonn}
A!2 = update(A!1, i, val)
\end{pythonn}
}
\end{minipage}

\begin{minipage}{0.33\textwidth}
{\small
\begin{cppp}
A_1[i] = val;
A_2 = A_1;
\end{cppp}
}
\end{minipage}

IMP Source & MPC Source & MOTION Code
\end{tabular}
\caption{MOTION Translation: Array Updates}
\label{tab:motion_translation_array_updates}
\end{table*}

% Phi nodes and FOR loops
Phi nodes are broken into two components: the ``false'' branch which assigns the initial value of the phi node and the ``true'' branch which assigns the phi node's backedge.  The assignment of the false branch occurs right before the Phi node's enclosing loop.  Inside of the Phi node's enclosing loop, a C++ \texttt{if} statement is inserted to only assign the true branch of the phi node after the first iteration.  MPC FOR loops are converted to C++ FOR loops which iterate the loop counter over the specified range.  The loop counter is initialized before any phi nodes in the loop as they could rely on it; there is no other special handling of FOR loops.

\begin{table*}
\begin{tabular}{cc}
\begin{minipage}{0.5\textwidth}
{\small
\begin{pythonn}
for i in range(N):
   tmp = PHI(arr[i], val!0)
	 ...
\end{pythonn}
}
\end{minipage}

\begin{minipage}{0.5\textwidth}
{\small
\begin{cppp}
_MPC_PLAINTEXT_i = 0;
tmp = arr[_MPC_PLAINTEXT_i];
for (; _MPC_PLAINTEXT_i < _MPC_PLAINTEXT_N; _MPC_PLAINTEXT_i++) {
   if (_MPC_PLAINTEXT_i != 0) {
      tmp = val_0;
	 }
	 ...
}
\end{cppp}
}
\end{minipage}

MPC Source & MOTION Code
\end{tabular}
\caption{MOTION Translation: FOR loop with Phi nodes}
\label{tab:motion_translation_for_loop}
\end{table*}

% SIMD representation
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional \texttt{std::vector}s in row-major order.  Whenever a SIMD value is used in an expression, the utility function \texttt{vectorized\_access()} takes the unvectorized representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value.  Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression.  Therefore the translation of an expression containing SIMD values is identical to that of expressions without SIMD values once vectorized accesses are inserted.

Similarly, the \texttt{vectorized\_assign()} function assigns a (potentially SIMD) value to a slice of a vectorized array.  This operation cannot be done with a simple subscript as SIMD assignments will update a range of values in the non-vectorized representation.
\ben{Is the next sentence needed?  They can always check the implementation of vectorized\_assign() if they want to see specifics.}
The function takes the array being assigned to as a reference to update it in place so that any existing values are not lost.  SIMD values are never directly assigned to non-SIMD values.

\begin{table*}
\begin{tabular}{cc}
\begin{minipage}{0.5\textwidth}
{\small
\begin{pythonn}
sum!4[I] = ADD_SIMD(sum!3[I], p[I, j])
\end{pythonn}
}
\end{minipage}

\begin{minipage}{0.5\textwidth}
{\small
% Ben: This line is too long but I don't know how to break it well
\begin{cppp}
vectorized_assign(sum_4, {_MPC_PLAINTEXT_N}, {true}, {}, vectorized_access(sum_3, {_MPC_PLAINTEXT_N}, {true}, {}) + vectorized_access(p, {_MPC_PLAINTEXT_N, _MPC_PLAINTEXT_D}, {true, false}, {_MPC_PLAINTEXT_j}));
\end{cppp}
}
\end{minipage}

MPC Source & MOTION Code
\end{tabular}
\caption{MOTION Translation: Assignment to SIMD value}
\label{tab:motion_translation_simd_assignment}
\end{table*}

% vectorized_update()
\ben{Technically this is semantically the same as non-vectorized update but we use a utility function to combine the steps into a single C++ statement - should it still be included?}
Updating SIMD arrays is also implemented differently from updating non-vectorized arrays.  Instead of separating the array update from the assignment of the new array, these steps are combined with the \texttt{vectorized\_update()} utility function.  This function operates identically to \texttt{vectorized\_assign()}, however it additionally returns the array after the assignment occurs.  This value is then used for the assignment to the new variable.

% raise_dim() implementation
\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...}
Raising the dimensions of an expression uses the \texttt{lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the \texttt{lift()} function which occurs when we are raising an array.  This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is).

% Ben (note as comment since tables aren't fixed placement): I only include the translation of the expression itself (instead of an assignment) since the assignment includes the extra junk from a vectorized_assign.  Also none of the raise_dims in biometric use the dimension index which I think is the most interesting/important part of the lift expression translation so I made up a toy example.
\begin{table*}
\begin{tabular}{cc}
\begin{minipage}{0.5\textwidth}
{\small
\begin{pythonn}
raise_dim(i + j, (i:N, j:M))
\end{pythonn}
}
\end{minipage}

\begin{minipage}{0.5\textwidth}
{\small
\begin{cppp}
lift(std::function([&](const std::vector<std::uint32_t> &idxs){return idxs[0] + idxs[1];}), {_MPC_PLAINTEXT_N, _MPX_PLAINTEXT_M})
\end{cppp}
}
\end{minipage}

MPC Source & MOTION Code
\end{tabular}
\caption{MOTION Translation: Raising dimensions}
\label{tab:motion_translation_raise_dim}
\end{table*}

% drop_dim() implementation
Dropping dimensions use the \texttt{drop\_dim()} and \texttt{drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.
\ben{drop\_dim() taking the true branch of phi node values doesn't need to be covered here, right?}

% Shared/plaintext conversions
\ben{Ishaq or Vassilis: please check over this paragraph... I don't want to make some mistake about MOTION's capabilities.}
Currently, our compiler only supports the \texttt{Bmr} and \texttt{BooleanGMW} protocols as MOTION does not implement all operations for other protocols \ben{Should that go here?}.  MOTION does not support publicly-known constants for these protocols, so all conversions from plaintext values to shares are performed by providing the plaintext value as a shared input from party 0.  Due to this limitation, our translation to MOTION code attempts to minimize the number of conversions into a plaintext value.  This is accomplished by creating a shared copy of each plaintext variable and updating that copy in lock-step with the plaintext variable.  Since variables are often initialized to a common constant value (e.g. 0), this approach decreases the number of input gates by only creating a shared input for each constant.  Loop counters must still be converted to a shared value on each iteration that they are used, however we only output this conversion when necessary to prevent loops where the counter is not used in a shared expression from increasing the number of input gates.

% Tons of copies != runtime costs
Due to the SSA translation phase as well as the various conversions to and from SIMD values which our utility functions perform, our generated MOTION code often includes many copies of arrays and scalar values.  These copies do not incur a runtime cost as the arrays simply hold pointers to the underlying shares, so no new shares or gates are created as a result of this copying. \ben{Is there anything else to say here?  Should this be placed somewhere else?}
